---
title: "BST260 Final Project: Understanding COVID-19 Moratality Rates of Nursing Homes in the US"
author: "Stella Nam"
output:
  pdf_document: default
  html_document: default
---

# Introduction

Over the past three years, the COVID-19 pandemic has affected the lives of millions worldwide. Although the global burden exists everywhere, the negative impact it has had on the senior population, especially those individuals residing in nursing homes, has been detrimental$^1$. It is critical to understand how and why so many senior residents in these homes lost their lives to the pandemic in order to implement prevention measures for this on-going COVID-19 pandemic and for future outbreaks. This project investigated the effect of quality of nursing homes on the number of COVID-19-related deaths in the US while controlling for geographic and socioeconomic factors. The quality of the residence was derived from the Centers for Medicare & Medicaid Services (CMS) 5-star quality rating system, a standardized measure created by the US federal agency$^2$.

The Nursing Home COVID-19 database in the US has been released on May 26, 2020. Since then, the database has been updated weekly by the CMS$^3$. The data retained in this data set is from June 1, 2020 to January 31, 2021 (8-month period) with information on nursing homes from all states except Guam and Puerto Rico. Over the months of data collection, the data set has 99,314 observations of 24 variables related to the nursing home, the local COVID-19 situation, and geographic determinants. 

The main outcome of the data set, mortality date in nursing homes, as well as variables that are directly related to the main outcome (COVOD-19 cases and non-COVID-related death cases in nursing home residents) were visualized to further understand the data set. The changes in COVID-19 cases per month seems to be following a trend proportional to the overall death rate in nursing homes. Additionally, different states reported varying levels of cases and deaths, but most states seem to be following the same trends over time (Fig. 1a, 1b). Therefore, it was not necessary to study a month-by-month difference in the death rates since most states appeared to follow the same trends. Rather, examining the overall difference during the entire 8-month study period to compare the state-by-state differences in nursing homes seems like a more appropriate approach for this data set. Additionally, the number of non-COVID-related deaths in the nursing homes tend to stay relatively constant over the study period (Fig. 1c). Therefore, using the overall death rate in nursing homes was deemed as an accurate method for studying COVID-related mortality rates in nursing homes without further data reshaping for specifically the main outcome. 

Since the original data set has information on each nursing home with each month being an entry, the data needs to be extensive reshaped to fit the research questions of this project. The month-by-month comparison needs to be averaged together as well as all the nursing home data by state. Moreover, certain predictors that do not align with the research question or cannot be properly analysed using averaged data were omitted. After data wrangling, a Poisson regression model is employed to estimate the COVID-19 mortality rates in nursing homes from our data. To account for overdispersion, an extension of the Poisson model, the Negative Binomial regression, was used to prevent overconfident estimates. No further modifications were made to the data set to perform a Negative Binomial regression because the data was already in counts and rate form. 

```{r, echo = F, message= F}
library(haven)
library(dplyr)
library(tidyr)
library(dslabs)
library(ggplot2)
library(ggpubr)
# the data set:
data <- read_dta('/Users/andy/Downloads/260projectfinal.dta')

########## START OF DATA WRANGLING ################

# remove all location variables except state
data2 <- data[ -c(1:3, 5:6) ]

# subset data into different months when data collected
month_0 = data2 |> filter(month_final == 0) |> rename("month_0" = "month_final")
month_1 = data2 |> filter(month_final == 1) |> rename("month_1" = "month_final")
month_2 = data2 |> filter(month_final == 2) |> rename("month_2" = "month_final")
month_3 = data2 |> filter(month_final == 3) |> rename("month_3" = "month_final")
month_4 = data2 |> filter(month_final == 4) |> rename("month_4" = "month_final")
month_5 = data2 |> filter(month_final == 5) |> rename("month_5" = "month_final")
month_6 = data2 |> filter(month_final == 6) |> rename("month_6" = "month_final")
month_7 = data2 |> filter(month_final == 7) |> rename("month_7" = "month_final")

# merge data for same states-- every variable represents the average value for the respective state
avg_state <- function(month_i) {
  month_i <- month_i[-c(17)]
  month_i |> group_by(providerstate) |> 
    summarize(RES_MonthlyconfC19cases = mean(RES_MonthlyconfC19cases, na.rm = TRUE),
              S_STAFF_ANY = mean(S_STAFF_ANY, na.rm = TRUE),
              S_PPE_ANY = mean(S_PPE_ANY, na.rm = TRUE),
              RES_Monthlyalldeaths_excptCOV = mean(RES_Monthlyalldeaths_excptCOV, na.rm = TRUE),
              overall_rating = mean(overall_rating, na.rm = TRUE),
              paymcaid_num = mean(paymcaid_num, na.rm = TRUE),
              pctwhite_2011pN = mean(pctwhite_2011pN, na.rm = TRUE),
              vulnerabilityindex_county = mean(vulnerabilityindex_county, na.rm = TRUE),
              CBSACode = mean(as.numeric(CBSACode), na.rm = TRUE),
              casesper100k_CBSA = mean(casesper100k_CBSA, na.rm = TRUE),
              deathsper100k_CBSA = mean(deathsper100k_CBSA, na.rm = TRUE),
              NH_deathsper100k = mean(NH_deathsper100k, na.rm = TRUE),
              ownership_recode = mean(ownership_recode, na.rm = TRUE),
              facility_size = mean(facility_size, na.rm = TRUE),
              urban_rural = mean(urban_rural, na.rm = TRUE),
              lnCBSAdeathRate = mean(lnCBSAdeathRate, na.rm = TRUE),
              lnCBSAcaseRate = mean(lnCBSAcaseRate, na.rm = TRUE))
}


# apply function and wrangle these 3 variables (NH_deathsper100k, RES_MonthlyconfC19cases, RES_Monthlyalldeaths_excptCOV) so that we can study their trends 
month_0 <- month_0 |> avg_state() |> mutate(month_0 = 0) |> 
  rename("RES_MonthlyconfC19cases_m0" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m0" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m0" = "NH_deathsper100k") 
month_1 <- month_1 |> avg_state() |> mutate(month_0 = 1) |> 
  rename("RES_MonthlyconfC19cases_m1" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m1" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m1" = "NH_deathsper100k") 
month_2 <- month_2 |> avg_state() |> mutate(month_0 = 2) |> 
  rename("RES_MonthlyconfC19cases_m2" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m2" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m2" = "NH_deathsper100k") 
month_3 <- month_3 |> avg_state() |> mutate(month_0 = 3) |> 
  rename("RES_MonthlyconfC19cases_m3" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m3" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m3" = "NH_deathsper100k") 
month_4 <- month_4 |> avg_state() |> mutate(month_0 = 4) |> 
  rename("RES_MonthlyconfC19cases_m4" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m4" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m4" = "NH_deathsper100k") 
month_5 <- month_5 |> avg_state() |> mutate(month_0 = 5) |> 
  rename("RES_MonthlyconfC19cases_m5" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m5" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m5" = "NH_deathsper100k") 
month_6 <- month_6 |> avg_state() |> mutate(month_0 = 6) |> 
  rename("RES_MonthlyconfC19cases_m6" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m6" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m6" = "NH_deathsper100k") 
month_7 <- month_7 |> avg_state() |> mutate(month_0 = 7) |> 
  rename("RES_MonthlyconfC19cases_m7" = "RES_MonthlyconfC19cases", 
         "RES_Monthlyalldeaths_excptCOV_m7" = "RES_Monthlyalldeaths_excptCOV",
         "NH_deathsper100k_m7" = "NH_deathsper100k") 


# ----------------------START OF EDA------------------------
# for some EDA, lets see if there's any intra- or inter-state trends in the covid deaths or cases
##monthly death rate for senior home residents 
month0 <- month_0 |> rename("0" = "NH_deathsper100k_m0") |>
  select(providerstate, "0")
month1 <- month_1 |> rename("1" = "NH_deathsper100k_m1") |>
  select(providerstate, "1") 
month2 <- month_2 |> rename("2" = "NH_deathsper100k_m2") |>
  select(providerstate, "2") 
month3 <- month_3 |> rename("3" = "NH_deathsper100k_m3") |>
  select(providerstate, "3") 
month4 <- month_4 |> rename("4" = "NH_deathsper100k_m4") |>
  select(providerstate, "4") 
month5 <- month_5 |> rename("5" = "NH_deathsper100k_m5") |>
  select(providerstate, "5")
month6 <- month_6 |> rename("6" = "NH_deathsper100k_m6") |>
  select(providerstate, "6") 
month7 <- month_7 |> rename("7" = "NH_deathsper100k_m7") |>
  select(providerstate, "7") 

total <- left_join(month0, month1, by="providerstate")
total <- left_join(total, month2, by="providerstate")
total <- left_join(total, month3, by="providerstate")
total <- left_join(total, month4, by="providerstate")
total <- left_join(total, month5, by="providerstate")
total <- left_join(total, month6, by="providerstate")
total <- left_join(total, month7, by="providerstate")

new_total <- total |> 
  pivot_longer("0":"7", names_to = "month", values_to = "res_death")
```

**Figure 1:**

```{r, echo=F}
# graph cases of residents:
fig1a <- new_total |> ggplot(aes(x = month, y= res_death)) +
  geom_line(aes(x = as.numeric(month), y= res_death, color = factor(providerstate))) +
  labs(x = "Month", y = "Monthly Avg Death Rate", 
              title = "Death Rate by State") +
  theme(legend.position = "none")
```
```{r, echo =F}
##monthly covid cases for senior home residents 
month0 <- month_0 |> rename("0" = "RES_MonthlyconfC19cases_m0") |>
  select(providerstate, "0")
month1 <- month_1 |> rename("1" = "RES_MonthlyconfC19cases_m1") |>
  select(providerstate, "1") 
month2 <- month_2 |> rename("2" = "RES_MonthlyconfC19cases_m2") |>
  select(providerstate, "2") 
month3 <- month_3 |> rename("3" = "RES_MonthlyconfC19cases_m3") |>
  select(providerstate, "3") 
month4 <- month_4 |> rename("4" = "RES_MonthlyconfC19cases_m4") |>
  select(providerstate, "4") 
month5 <- month_5 |> rename("5" = "RES_MonthlyconfC19cases_m5") |>
  select(providerstate, "5")
month6 <- month_6 |> rename("6" = "RES_MonthlyconfC19cases_m6") |>
  select(providerstate, "6") 
month7 <- month_7 |> rename("7" = "RES_MonthlyconfC19cases_m7") |>
  select(providerstate, "7") 

total <- left_join(month0, month1, by="providerstate")
total <- left_join(total, month2, by="providerstate")
total <- left_join(total, month3, by="providerstate")
total <- left_join(total, month4, by="providerstate")
total <- left_join(total, month5, by="providerstate")
total <- left_join(total, month6, by="providerstate")
total <- left_join(total, month7, by="providerstate")

new_total <- total |> 
  pivot_longer("0":"7", names_to = "month", values_to = "res_cases")

# graph cases of residents:
fig1b <- new_total |> ggplot(aes(x = month, y= res_cases)) +
  geom_line(aes(x = as.numeric(month), y= res_cases, color = factor(providerstate))) +
  labs(x = "Month", y = "Monthly Avg Covid Cases", 
              title = "Covid cases by State") +
  theme(legend.position = "none")

```
```{r, echo =F, fig.height = 5, fig.width = 10}
##monthly non-covid-related death cases for senior home residents 
month0 <- month_0 |> rename("0" = "RES_Monthlyalldeaths_excptCOV_m0") |>
  select(providerstate, "0")
month1 <- month_1 |> rename("1" = "RES_Monthlyalldeaths_excptCOV_m1") |>
  select(providerstate, "1") 
month2 <- month_2 |> rename("2" = "RES_Monthlyalldeaths_excptCOV_m2") |>
  select(providerstate, "2") 
month3 <- month_3 |> rename("3" = "RES_Monthlyalldeaths_excptCOV_m3") |>
  select(providerstate, "3") 
month4 <- month_4 |> rename("4" = "RES_Monthlyalldeaths_excptCOV_m4") |>
  select(providerstate, "4") 
month5 <- month_5 |> rename("5" = "RES_Monthlyalldeaths_excptCOV_m5") |>
  select(providerstate, "5")
month6 <- month_6 |> rename("6" = "RES_Monthlyalldeaths_excptCOV_m6") |>
  select(providerstate, "6") 
month7 <- month_7 |> rename("7" = "RES_Monthlyalldeaths_excptCOV_m7") |>
  select(providerstate, "7") 

total <- left_join(month0, month1, by="providerstate")
total <- left_join(total, month2, by="providerstate")
total <- left_join(total, month3, by="providerstate")
total <- left_join(total, month4, by="providerstate")
total <- left_join(total, month5, by="providerstate")
total <- left_join(total, month6, by="providerstate")
total <- left_join(total, month7, by="providerstate")

new_total <- total |> 
  pivot_longer("0":"7", names_to = "month", values_to = "res_nonCovidDeath")

# graph cases of residents:
fig1c <- new_total |> ggplot(aes(x = month, y= res_nonCovidDeath)) +
  geom_line(aes(x = as.numeric(month), y= res_nonCovidDeath, color = factor(providerstate))) +
  labs(x = "Month", y = "Monthly Avg Non-Covid Deaths", 
              title = "Non-Covid Deaths by State", color = "State") +
  theme(legend.key.size = unit(.1, 'cm'), #change legend key size
        legend.key.height = unit(.2, 'cm'), #change legend key height
        legend.key.width = unit(.1, 'cm'), #change legend key width
        legend.title = element_text(size=7), #change legend title font size
        legend.text = element_text(size=5))

ggarrange(fig1a, fig1b, fig1c, 
          labels = c("A", "B", "C"),
          ncol = 2, nrow = 2)

# -----------------------END OF EDA------------------
```

# Results (Route 1)

***Data wrangling***

The original data set was organized such that each row entry was for a nursing home in the US at each month of the study period. Therefore, there were 99,314 row entries of 12415 different senior homes across the US (Table 1). This data was not tidy. Additionally, it was not useful for studying state-by-state differences over the entire study period. Therefore, all location-related variables except the state variable was removed. All the data was grouped for all the data by each month and by state. Thus, there were 8 row entries for each state with each row representing a different month in the study period. Upon analysis of the variations between the study period (refer to Introduction), we decided to study patterns in mortality rates over the entire study period rather than at different time points. Hence, the data was further reshaped to display the overall averaged values for each state over the entire study period. All data from the Core Based Statistical Area (CBSA) were removed as they were not relevant for a state-by-state averaged study since CBSA data would be more useful for indidivual, within-state research questions. Therefore, the final data set for our analysis contained 51 row entires (one for each state), 1 variable for the main outcome (mortality rate), and 10 potential predictors of the main outcome (Table 2).

```{r, echo = F}
######### DATA WRANGLING CONTINUED ################

# now finish wrangling the data 
# find mean of all variables -- since we are doing overall study period approach
NH_deathsper100k <- data.frame(month_0$NH_deathsper100k_m0, month_1$NH_deathsper100k_m1, month_2$NH_deathsper100k_m2, month_3$NH_deathsper100k_m3, month_4$NH_deathsper100k_m4, month_5$NH_deathsper100k_m5, month_6$NH_deathsper100k_m6, month_7$NH_deathsper100k_m7)
NH_deathsper100k_mean <- rowMeans(NH_deathsper100k)

RES_MonthlyconfC19cases <- data.frame(month_0$RES_MonthlyconfC19cases_m0, month_1$RES_MonthlyconfC19cases_m1, month_2$RES_MonthlyconfC19cases_m2, month_3$RES_MonthlyconfC19cases_m3, month_4$RES_MonthlyconfC19cases_m4, month_5$RES_MonthlyconfC19cases_m5, month_6$RES_MonthlyconfC19cases_m6, month_7$RES_MonthlyconfC19cases_m7)
RES_MonthlyconfC19cases_mean <- rowMeans(RES_MonthlyconfC19cases)

RES_Monthlyalldeaths_excptCOV <- data.frame(month_0$RES_Monthlyalldeaths_excptCOV_m0, month_1$RES_Monthlyalldeaths_excptCOV_m1, month_2$RES_Monthlyalldeaths_excptCOV_m2, month_3$RES_Monthlyalldeaths_excptCOV_m3, month_4$RES_Monthlyalldeaths_excptCOV_m4, month_5$RES_Monthlyalldeaths_excptCOV_m5, month_6$RES_Monthlyalldeaths_excptCOV_m6, month_7$RES_Monthlyalldeaths_excptCOV_m7)
RES_Monthlyalldeaths_excptCOV_mean <- rowMeans(RES_Monthlyalldeaths_excptCOV)

S_STAFF_ANY <- data.frame(month_0$S_STAFF_ANY, month_1$S_STAFF_ANY, month_2$S_STAFF_ANY, month_3$S_STAFF_ANY, month_4$S_STAFF_ANY, month_5$S_STAFF_ANY, month_6$S_STAFF_ANY, month_7$S_STAFF_ANY)
S_STAFF_ANY_mean <- rowMeans(S_STAFF_ANY)

S_PPE_ANY <- data.frame(month_0$S_PPE_ANY, month_1$S_PPE_ANY, month_2$S_PPE_ANY, month_3$S_PPE_ANY, month_4$S_PPE_ANY, month_5$S_PPE_ANY, month_6$S_PPE_ANY, month_7$S_PPE_ANY)
S_PPE_ANY_mean <- rowMeans(S_PPE_ANY)

overall_rating <- data.frame(month_0$overall_rating, month_1$overall_rating, month_2$overall_rating, month_3$overall_rating, month_4$overall_rating, month_5$overall_rating, month_6$overall_rating, month_7$overall_rating)
overall_rating_mean <- rowMeans(overall_rating)

paymcaid_num <- data.frame(month_0$paymcaid_num, month_1$paymcaid_num, month_2$paymcaid_num, month_3$paymcaid_num, month_4$paymcaid_num, month_5$paymcaid_num, month_6$paymcaid_num, month_7$paymcaid_num)
paymcaid_num_mean <- rowMeans(paymcaid_num)

pctwhite_2011pN <- data.frame(month_0$pctwhite_2011pN, month_1$pctwhite_2011pN, month_2$pctwhite_2011pN, month_3$pctwhite_2011pN, month_4$pctwhite_2011pN, month_5$pctwhite_2011pN, month_6$pctwhite_2011pN, month_7$pctwhite_2011pN)
pctwhite_2011pN_mean <- rowMeans(pctwhite_2011pN)

vulnerabilityindex_county <- data.frame(month_0$vulnerabilityindex_county, month_1$vulnerabilityindex_county, month_2$vulnerabilityindex_county, month_3$vulnerabilityindex_county, month_4$vulnerabilityindex_county, month_5$vulnerabilityindex_county, month_6$vulnerabilityindex_county, month_7$vulnerabilityindex_county)
vulnerabilityindex_county_mean <- rowMeans(vulnerabilityindex_county)

CBSACode <- data.frame(month_0$CBSACode, month_1$CBSACode, month_2$CBSACode, month_3$CBSACode, month_4$CBSACode, month_5$CBSACode, month_6$CBSACode, month_7$CBSACode)
CBSACode_mean <- rowMeans(CBSACode)

casesper100k_CBSA <- data.frame(month_0$casesper100k_CBSA, month_1$casesper100k_CBSA, month_2$casesper100k_CBSA, month_3$casesper100k_CBSA, month_4$casesper100k_CBSA, month_5$casesper100k_CBSA, month_6$casesper100k_CBSA, month_7$casesper100k_CBSA)
casesper100k_CBSA_mean <- rowMeans(casesper100k_CBSA)

deathsper100k_CBSA <- data.frame(month_0$deathsper100k_CBSA, month_1$deathsper100k_CBSA, month_2$deathsper100k_CBSA, month_3$deathsper100k_CBSA, month_4$deathsper100k_CBSA, month_5$deathsper100k_CBSA, month_6$deathsper100k_CBSA, month_7$deathsper100k_CBSA)
deathsper100k_CBSA_mean <- rowMeans(deathsper100k_CBSA)

ownership_recode <- data.frame(month_0$ownership_recode, month_1$ownership_recode, month_2$ownership_recode, month_3$ownership_recode, month_4$ownership_recode, month_5$ownership_recode, month_6$ownership_recode, month_7$ownership_recode)
ownership_recode_mean <- rowMeans(ownership_recode)

facility_size <- data.frame(month_0$facility_size, month_1$facility_size, month_2$facility_size, month_3$facility_size, month_4$facility_size, month_5$facility_size, month_6$facility_size, month_7$facility_size)
facility_size_mean <- rowMeans(facility_size)

urban_rural <- data.frame(month_0$urban_rural, month_1$urban_rural, month_2$urban_rural, month_3$urban_rural, month_4$urban_rural, month_5$urban_rural, month_6$urban_rural, month_7$urban_rural)
urban_rural_mean <- rowMeans(urban_rural)

lnCBSAdeathRate <- data.frame(month_0$lnCBSAdeathRate, month_1$lnCBSAdeathRate, month_2$lnCBSAdeathRate, month_3$lnCBSAdeathRate, month_4$lnCBSAdeathRate, month_5$lnCBSAdeathRate, month_6$lnCBSAdeathRate, month_7$lnCBSAdeathRate)
lnCBSAdeathRate_mean <- rowMeans(lnCBSAdeathRate)

lnCBSAcaseRate <- data.frame(month_0$lnCBSAcaseRate, month_1$lnCBSAcaseRate, month_2$lnCBSAcaseRate, month_3$lnCBSAcaseRate, month_4$lnCBSAcaseRate, month_5$lnCBSAcaseRate, month_6$lnCBSAcaseRate, month_7$lnCBSAcaseRate)
lnCBSAcaseRate_mean <- rowMeans(lnCBSAcaseRate)

# now, merge all the averaged data together
# exclude CBSA rate data -- not necessary for research questions 
data_final <- data.frame("state" = month_0$providerstate, NH_deathsper100k_mean, RES_MonthlyconfC19cases_mean, RES_Monthlyalldeaths_excptCOV_mean, S_STAFF_ANY_mean, S_PPE_ANY_mean, overall_rating_mean, paymcaid_num_mean, pctwhite_2011pN_mean, vulnerabilityindex_county_mean, facility_size_mean, urban_rural_mean)
```

**Table 1: Original Data Set**

```{r, echo = F}
head(data)
```

**Table 2: Wrangled Data Set only including the first 6 variables**
```{r, echo =F}
head(as_tibble(data_final))
```

***Analysis: Poisson Regression***

Overdispersion:

As a model specialized for count data, we employed a Poisson regression. Although the Poisson density roughly fits each of our potential target variables, they are overdispersed because the variance is greater than the mean (nursing home death rate: mean of 1110.511 deaths/100,000 residents and variance of 160259.1 deaths$^2$/100,000 residents$^2$ over the 8-month study period (Fig. 2). The presence of overdispersion suggest the use of a Poisson model extension, rather than the standard Poisson model itself. With a lack of zero-inflated data, we employ a Negative Binomial regression, which should account for overdispersion, to make sure we do not have overconfident estimates. For these data, we do not need to account for lag as they are all averaged over the same time period for each predictor and outcome variable, respectively.

**Figure 2: Average Covid-19 Death Rate Density Distribution**

```{r, echo =F, fig.height = 3, fig.width = 4}
############## DATA ANAYLSIS - POISSON DISTRIBUTION ##########
###### CHECK OVER DISPERSION 

# death rate in nursing homes study period 
hist(data_final$NH_deathsper100k_mean, freq = F, ylim = c(0, 0.012), main = NULL, xlab = "Average Death Rate", ylab = "Density")
lines(as.integer(min(data_final$NH_deathsper100k_mean)):as.integer(max(data_final$NH_deathsper100k_mean)), dpois(as.integer(min(data_final$NH_deathsper100k_mean)):as.integer(max(data_final$NH_deathsper100k_mean)), lambda = mean(data_final$NH_deathsper100k_mean)))
```

```{r, echo = F, eval =F}
# mean and variance calculations to check for overdispersion
print(mean(data_final$NH_deathsper100k_mean))
print(var(data_final$NH_deathsper100k_mean))
```

Correlation:

The main outcome is death rate (NH_deathsper100k_mean) and the main predictor is nursing home quality (overall_rating_mean). We want to understand how correlated the two variables are in order to get a better grasp for how nursing home quality is related to COVID-19 death rates. 

Based on plots below, there is a lot of variation between the main predictor and outcome from the large scatter (Fig. 3). However, there is a downward trend where the higher the nursing home rate, the lower the morality rate among the residents. This is confirmed by the correlation coefficient (R = -0.4471934), which suggests that the variables are rougly negatively correlated. This implies that there is some relationship that should be further studied along with other potential predictors of mortality rate in these nursing homes. 

**Figure 3:**

```{r, echo = F, message = F, fig.height = 3, fig.width = 10}
######## VARIABLE SELECTION AND MODEL FITS 

# CHECK CORRELATIONS
# how correlated are death rate and nursing home ratings?
fig3a <- data_final |> 
  ggplot(aes(overall_rating_mean, NH_deathsper100k_mean)) + 
  geom_point() +
  geom_smooth() +
  labs(x = "Average Nursing Home Ratings", y = "Average Mortality Rate", title = "Average Mortality Rate by Nursing Home Quality")

fig3b <- data_final |> mutate(rating_strata = factor(round(overall_rating_mean, digits = 0.5))) |> 
  ggplot(aes(rating_strata, NH_deathsper100k_mean)) + 
  geom_boxplot() + 
  geom_point() +
  labs(x = "Average Nursing Home Ratings", y = "Average Mortality Rate", title = "Boxplot of Avg Mortality Rate by Nursing Home Quality")

ggarrange(fig3a, fig3b, 
          labels = c("A", "B"),
          ncol = 2, nrow = 1)

```

```{r, echo = F, eval = F}
data_final |> summarize(r = cor(NH_deathsper100k_mean, overall_rating_mean)) |> pull(r)
# -0.4471934 -- variables are negatively correlated, but not extremely strong relationship. Will control for other variables 
```

Model Fit:

The negative binomial model with just the main predictor (nursing home quality) with mortality rate as the outcome had a relatively high AIC value of 763.81 despite the predictor being significant (p-value < 0.05). We then adjusted for the number of COVID-19 cases in the nursing home since there was some variability observed from the exploratory data analysis (refer to Introduction). Additionally, the COVID-19 cases in nursing home residents could be a potential confounder since it is could be associated with both the outcome (death rate) and main predictor (nursing home quality), but is not a downstream consequence of the main predictor. This decreased the AIC value to 756.01, thus suggesting an improved model fit for our data, however, the main predictor was no longer significant. Therefore, we checked for effect modification of COVID-19 cases on the main predictor effect on the main outcome. We observed effect modification with a improved model fit upon including the interaction term. Similarly, we controlled for nursing home size, staff shortages, PPE shortages for same reason of potential confounding while checking for effect modification. As a result, we had an improved model with an AIC value of 729.36.

Moreover, we controlled for socioeconomic factors like percentage of white residents in the facility and percentage of Medicaid residents in the facility since they can be determinants of death rate. However, we did not control for whether the average nursing home was more urban or rural since the variable in the source data set was binary (0 for rural and 1 for urban). Our data was averaged over a large sample size for this binary operator and thus was not deemed to be effective at suggesting whether the homes were more rural or urban on average. Similarly, we did not adjust for CDC social vulnerability index for the same reason of being a binary predictor averaged over a large sample size. (Note: detailed step-by-step analysis of the model fit can be found in the Appendix.)

```{r, echo = F, warning = F, eval = F}
# MODELING
# poisson fit general only with main predictor 
pois_general <- data_final %>% glm(formula = NH_deathsper100k_mean ~ overall_rating_mean, family=poisson(), data=.)
summary(pois_general)$coefficients
summary(pois_general)$aic

# neg bin fit general only with main predictor 
neg_bin <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean, data=.)
summary(neg_bin)$coefficients
summary(neg_bin)$aic # AIC = 763.81

#---------------------#

# include monthly covid cases of residents in facility 
neg_bin_2 <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean, data=.)
summary(neg_bin_2)$coefficients
summary(neg_bin_2)$aic
# AIC = 756.01 (decreases), but main predictor no longer significant

neg_bin_2_EM <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean, data=.)
summary(neg_bin_2_EM)$coefficients
summary(neg_bin_2_EM)$aic
# There is effect modification since p-value of interaction term <0.05
# AIC = 747.35 (decreases), AND main predictor NOW significant again 

# include facility size 
neg_bin_3 <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean, data=.)
summary(neg_bin_3)$coefficients
summary(neg_bin_3)$aic
# AIC = 731.4 (decreases) and all significant co-variates

neg_bin_3_EM <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean + facility_size_mean*overall_rating_mean, data=.)
summary(neg_bin_3_EM)$coefficients
# No effect modification of facility size on overall rating 
# thus, keep neg_bin_3

# include shortages
neg_bin_4a <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean + S_STAFF_ANY_mean + S_PPE_ANY_mean, data=.)
summary(neg_bin_4a)$coefficients
summary(neg_bin_4a)$aic
# AIC = 731.33 (almost same) and S_PPE_ANY_mean not significant 

neg_bin_4b <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean + S_STAFF_ANY_mean, data=.)
summary(neg_bin_4b)$coefficients
summary(neg_bin_4b)$aic
# AIC = 729.36 (decreases) 

# include SES factors
neg_bin_5a <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean + S_STAFF_ANY_mean + pctwhite_2011pN_mean + paymcaid_num_mean, data=.)
summary(neg_bin_5a)$coefficients
summary(neg_bin_5a)$aic
# AIC = 685.14 (decreases) and pctwhite_2011pN_mean not significant 

neg_bin_5b <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean + S_STAFF_ANY_mean + paymcaid_num_mean, data=.)
summary(neg_bin_5b)$coefficients
summary(neg_bin_5b)$aic 
# AIC = 685.11 (decreases slightly) and all covariates significant (p-value <0.05) # final model!! 
```


Therefore, we derive our final model (Table 3). $$log(E[Y_i = mortality rate | X_j] = \beta_0 + \beta_1X_1+…+\beta_pX_p$$

**Table 3: Fitted Negative Binomial regression model summary**

```{r, echo = F, warning =F}
neg_bin_5b <- data_final %>% MASS::glm.nb(formula = NH_deathsper100k_mean ~ overall_rating_mean + RES_MonthlyconfC19cases_mean + overall_rating_mean*RES_MonthlyconfC19cases_mean + facility_size_mean + S_STAFF_ANY_mean + paymcaid_num_mean, data=.)
summary(neg_bin_5b)$coefficients
```

Note that "estimate" is equivalent to a $\beta$ value.

With an AIC value of 685.11, the lowest value of all the other models we have explored, leads us to believe that it is the best-fit model of the models we have considered. Additionally, we find that the negative binomial regression has slightly higher standard errors and resulting test statistics compared to the standard Poisson regression. This is as we expected since it suggests that the standard Poisson regression may cause overconfidence in model evaluations, due to the property of E[Y] = Var(Y) being unsatisfied.

Furthermore, we can interpret the Negative Binomial regression supposing nursing home quality is our covariate of interest. For a 1-star increase in nursing home quality rating, a state has an estimated $e^{-1.51345882} = 0.2201472$ times the incident rate of mortality rate in nursing home residents compared to if a state did not have a 1-star increase in nursing home quality rating, on average, holding all other covariates (monthly average COVID-19 cases among residents, average facility size, average staff storage, and average Medicaid resident percentage) fixed. This was a significant relationship as well, suggesting that nursing home quality may be negatively associated with the incidence rate of COVID-19 deaths according to these data.

# Conclusion

Overall, our findings using the negative binomial regression model (an extension of the Poisson regression model) agree with previous literature and further elucidate the role nursing home quality on COVID-19 mortality rates in nursing homes across the US. Additionally, other factors such as the severity of the COVID-19 pandemic in the state and socioeconomic status can influence the mortality rate in senior homes. 

Throughout our model variations, we consistently observe a potential decrease in COVIOD-19 mortality rates on average following an increase in nursing home quality, according to these data. This, combined with the observation that an increase in staff shortages is associated with a potential increase in mortality rate suggests that the lack of access to care personnel such as nurses and clinicians could be potentially driving worse health outcomes. Interestingly, despite the racial composition of the nursing homes not significantly altering mortality rates, a potential decrease in mortality rate was observed as the percentage of medicaid residents increased. This further confirms the impact of socioeconomic influences on the senior population in the US. 

If given more time, I would have liked to explore differences in healthcare access and training of the staff in the nursing homes. These factors could be important predictors for understanding the morality rates. Additionally, updating the data set to contain up-to-date data on the mortality rates would have been another future direction of interest given that the COVID-19 pandemic is still an ongoing outbreak. Understanding how vaccines and the differences in vaccine roll out in the different states affected the senior population, specifically those living in nursing homes, could have been interesting to investigate.

# References

1. Cronin CJ, Evans WN. Nursing home quality, COVID-19 deaths, and excess mortality. *J Health Econ*. 2022;82:102592. doi:10.1016/j.jhealeco.2022.102592

2. Five-star quality rating system. CMS. https://www.cms.gov/medicare/provider-enrollment-and-certification/certificationandcomplianc/fsqrs. Published December 6, 2022. Accessed December 16, 2022. 

3. Gupta DD, Kelekar U, Turner SC, Sule AA, Jerman TG. Cov-19 deaths (monthly) among nursing home residents in the US: June-1, 2020 - January-31, 2021. openICPSR. https://www.openicpsr.org/openicpsr/project/147921/version/V1/view?path=%2Fopenicpsr. Published August 20, 2021. Accessed December 16, 2022. 

# Appendix


```{r ref.label=knitr::all_labels(), echo=TRUE, eval=T, warning=F}
```